---
title: "Creating Animated Race Bar Chart in R"
author: "Harry Aginta"
output:
  html_document:
    code_download: true
    df_print: paged
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 4
    number_sections: true
    code_folding: "show"
    theme: "cosmo"
    highlight: "monochrome"
  pdf_document: default
  html_notebook:
    code_folding: show
    highlight: monochrome
    number_sections: yes
    theme: cosmo
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: no
      smooth_scroll: no
  word_document: default
---

<style>
h1.title {font-size: 18pt; color: DarkBlue;} 
body, h1, h2, h3, h4 {font-family: "Palatino", serif;}
body {font-size: 12pt;}
/* Headers */
h1,h2,h3,h4,h5,h6{font-size: 14pt; color: #00008B;}
body {color: #333333;}
a, a:hover {color: #8B3A62;}
pre {font-size: 12px;}
</style>

This project uses GDP dataset downloaded from the WorldBank Database. The dataset can be accessed in [GitHub Repository](https://github.com/haginta/animated-bar-chart-in-R)


# Part A: Loading libraries and data pre-processing
## Libraries
```{r message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(janitor)
library(gganimate)
#install.packages("gifski")
library(gifski)
library(readr)
```

## Load GDP data
```{r}
gdp <- read_delim("gdp_ori.csv", 
    ";", escape_double = FALSE, trim_ws = TRUE)
```

```{r}
head(gdp)
```

```{r}
gdp
```

Drop non-country rows
```{r}
gdp_balance <- filter(gdp, !(country_code %in% c("ARB", "CEB", "CSS","EAP", "EAR", "EAS", "ECA", "ECS", "EMU", "EUU", "FCS", "HIC", "HPC", "IBD", "IBT", "IDA", "IDB", "IDX", "INX", "LAC", "LCN", "LDC", "LIC", "LMC", "LMY", "LTE", "MIC", "MEA", "MNA", "NAC", "OED", "OSS", "PST", "PSS", "PRE", "SSA", "SSF", "TEA", "TLA", "TMN", "TSA", "TSS", "UMC", "WLD")))
gdp_balance
```

Drop all rows with NA in "Region" column
```{r}
gdp_balance <- subset(gdp_balance, !is.na(Region))
gdp_balance
```

Transform the dataframe into long shape
```{r}
gdp_long <- gdp_balance %>% 
  mutate_at(vars(contains("yr")),as.numeric) %>% 
  gather(year,gdp,5:54)
```

## Save the dataframe that is ready to use
```{r}
write_csv(gdp_long, "gdp_ready.csv")
```

# Part B: Building the static plots using ggplot
## Import the ready-to-use data
```{r}
gdp_ready <- read_csv("gdp_ready.csv")
gdp_ready
```

```{r}
gdp_set <- gdp_ready %>%
  group_by(year) %>%
  mutate(rank = rank(-gdp),
         gdp_rel = gdp/gdp[rank==1],
         gdp_lbl = paste0(" ",round(gdp/1e9))) %>%
  group_by(country_name) %>% 
  filter(rank <=30) %>%
  ungroup()
```

```{r}
static_plot <- ggplot(gdp_set, aes(rank, group = country_name)) +  
  geom_tile(aes(y = gdp/2,
                height = gdp, fill = Region,
                width = 0.9), alpha = 0.8, color = NA) +
  geom_text(aes(y = 0, label = paste(country_name, " ")), vjust = 0.2, hjust = 1) +
  geom_text(aes(y=gdp,label = gdp_lbl, hjust=0)) +
  geom_text(aes(x=30, y=max(gdp) , label = as.factor(year)), vjust = 0.2, alpha = 0.5,  col = "gray", size = 20) +
  coord_flip(clip = "off", expand = FALSE) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_reverse() +
  scale_fill_discrete(guide = guide_legend(title.theme = element_text(
      size = 20), label.theme = element_text(size = 15))) +
  theme(axis.line=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        panel.background=element_blank(),
        panel.border=element_blank(),
        panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        panel.grid.major.x = element_line( size=.1, color="grey" ),
        panel.grid.minor.x = element_line( size=.1, color="grey" ),
        plot.title=element_text(size=25, hjust=0, face="bold", colour="black", vjust=-1),
        plot.subtitle=element_text(size=18, hjust=1, face="italic", color="grey"),
        plot.caption =element_text(size=14, hjust=1, face="italic", color="grey"),
        plot.background=element_blank(),
       plot.margin = margin(2,2, 2, 4, "cm"))
```

# Part C: Animating the static plot
```{r}
animated <- static_plot + transition_states(year,
transition_length = 3, state_length = 0, wrap = FALSE) +
  view_follow(fixed_x = TRUE)  +
  ease_aes('linear')+
    enter_fade()+
    exit_fade() +
  labs(title = 'Top 30 World GDP, 1970-2019',  
       subtitle  =  "GDP in Billions USD",
       caption  = "Data Source: World Bank Data")
```


## Render the animated chart to .gif file
```{r message=FALSE, warning=FALSE}
animate(animated, 150, fps = 5, end_pause = 30, width = 1500, height = 1000, 
        renderer = gifski_renderer("anim_gdp.gif"))
```


